Dialect: A Compact Line Notation Based on SMILES

### 2022-05-25 manuscript.tex
Reflows Conclusion.

Another advantage is lossless serialization and deserialization. The line notation InChI might appear to compete with Dialect in this sense. Although InChI can be read and written, the InChI Technical manual notes that InChI's value lies in molecular identification. The authors of InChI have repeatedly noted that InChI is not a solution to the molecular serialization problem. The reason presumably lies with the fact that both serialization and deserialization require intimate knowledge of the InChI canonicalization algorithm. To date, no third-party reader or writer of InChI has ever been released.

Despite its information density, Dialect can faithfully encode and decode most of what chemists would consider \enquote{organic molecules.} This is evidenced by the widespread presence of Dialect strings within large, public-facing databases such as PubChem, ChEMBL, ChEBI, and eMolecules.

@@

Broader expansion of Dialect's capabilities could be possible through \textit{metaformats}. A metaformat embeds one or more Dialect strings within a surrounding serialization format.For example, two- or three-dimensional coordinates could be associated with each atom through a metaformat that includes a dictionary mapping atomic index to coordinate. Collections of atoms or bonds could likewise be encoded to replicate the enhanced stereochemistry features of other formats. And so on. However, the utility of such extensions should be weighed against Dialect's main value proposition: high information density. An application attempting to use a verbose metaformat may benefit from adopting better-suited format instead.

@@

The availability of a minimal yet highly functional, fully-specified core language offers many opportunities to improve data quality. One of the most important will be a reference implementation, the design and source code of which will be discussed later. A freely-available reference implementation in turn makes automated validation suites possible. These suites can improve data quality by reporting syntax and semantic differences among implementations, preferably before release. A reference implementation taken together with the guidelines for readers and writers in this paper should make it possible to write software that conforms to a very high level of precision, regardless of programming language or paradigm. Given verified implementations, performance optimizations can be considered. The existence of a core language specification should also aid standardization efforts, either for Dialect itself, or SMILES. Finally, the development of better line notations is only possible given a thorough understanding of the scope and limitations of existing options. Here, metaformats could offer a bridge from the present to the future.

@@

Dialect's syntax is described in detail through both graphical diagrams and a formal grammar. The latter method concisely summarizes the complete set of strings that could be considered valid Dialect representations. The formal grammar was deliberately structured to be used either directly with an automated parser generator or with a hand-written recursive-descent parser.

>>>>

Another advantage is lossless serialization and deserialization. The line notation InChI might appear to compete with Dialect in this sense. The authors of InChI have noted that InChI is not a solution to the molecular serialization problem but rather an identifier.\cite{heller:2015} The reason presumably lies with the fact that both serialization and deserialization require intimate knowledge of the InChI canonicalization algorithm. To date, no third-party reader or writer of InChI has ever been released.

Despite its information density, Dialect can faithfully encode and decode most of what chemists would consider \enquote{organic molecules.} This is evidenced by the widespread presence of Dialect strings within large, public-facing databases such as PubChem,\cite{kim:2016} ChEMBL,\cite{gaulton:2012} ChEBI,\cite{degtyarenko:2007} and others.

@@

Broader expansion of Dialect's capabilities could be possible through \textit{metaformats}. A metaformat embeds one or more Dialect strings within a surrounding serialization format. For example, two- or three-dimensional coordinates could be associated with each atom through a metaformat that includes a dictionary mapping atomic index to coordinate. Collections of atoms or bonds could likewise be encoded to replicate the enhanced stereochemistry features of other formats. And so on. However, the utility of such extensions should be weighed against Dialect's main value proposition: high information density. An application attempting to use a verbose metaformat may benefit from adopting better-suited format instead.

@@

The availability of a minimal yet highly functional, fully-specified core language offers many opportunities to improve data quality. One of the most important will be an open reference implementation, the design and source code of which will be discussed later. A freely-available reference implementation in turn makes automated validation suites possible. These suites can improve data quality by reporting syntax and semantic differences among implementations, preferably before release. A reference implementation taken together with the guidelines for readers and writers in this paper should make it possible to write software that conforms to a very high level of precision, regardless of programming language or paradigm. Given verified implementations, performance optimizations can be considered. The existence of a core language specification should also aid standardization efforts, either for Dialect itself, or SMILES. Finally, the development of better line notations is only possible given a thorough understanding of the scope and limitations of existing options. Here, metaformats could offer a bridge from the present to the future.

@@

Dialect's syntax is described in detail through a formal grammar. This method concisely summarizes the complete set of strings that could be considered syntactically valid Dialect representations. The formal grammar was deliberately structured to be used either directly with an automated parser generator or with a hand-written recursive-descent parser.
